---
title: "Effects of Housing Problems and Home Life on Overall Health"
Author: "Brett Vetter, Dr. Chris Merkord"
Presentation ID: 4994
Department: Biosciences Department, Minnesota State University Moorhead, 1104 7th Avenue South, Moorhead, MN 56563 USA
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
```

## Abstract 

  Often times when people think of public health, the home you live in does not come to mind. But many people all over the country and in Minnesota have severe housing problems. Homes with severe housing problems are households with at least one of four housing problems: overcrowding, high housing costs, lack of kitchen facilities, or lack of plumbing facilities. These things affect your everyday life and tend to be ignored when it comes to your overall health. People tend to not correlate living in poverty or our mental health at home with our overall health either. So the question I have is, do severe housing problems, living in poverty growing up and living with untreated poor mental health have an effect on your overall health and your life expectancy?  I will be using data from the Minnesota Health Outcomes data set, from countyhealthrankings.org. The data set consists of variables pertaining to health outcomes from the year 2020, specifically looking at the counties in Minnesota. 

## Introduction

 My exploratory data analysis will look at the correlation between % of severe housing problems, average number of mentally unhealthy days and % of children in poverty with life expectancy and % of fair or poor health among counties in Minnesota. I will be looking at the correlation some unaddressed issues I mentioned have on our overall health, along with the correlation other more addressed issues have on our overall health. I believe the results of this analysis will help medical  and public health workers better focus on simple health factors like where people eat, sleep and spend most of their time to help diagnose and more efficiently treat people’s health issues. When treating patients for health issues we need to pay more attention to the factors that may be the root to their issues and correcting those factors within our communities. 

## Methods

### Data Acquisition

I found my data on countyhealthrankings.org (U of Wisconsin 2020). I specifically looked at the Minnesota county health rankings. There I found and chose the excel sheet "2020 Minnesota Data" and downloaded it. After I downloaded the excel sheet, I saved the data set to my data folder I created.

### Data Description

My data consists of 90 rows, each row representing a different county in Minnesota. The variables I chose to use from the data set were:
Deaths - Numerical Discrete 
% Housing problems - Numerical Continuous 
% Fair or poor health - Numerical Continuous
Life expectancy - Numerical Continuous 
Average number of mentally unhealthy days - Numerical Continuous
% of children in poverty - Numerical Continuous

### Data Preparation

For this data analysis I used Rstudio version 1.3.1093 (RStudio Team 2020) which is in R version 4.0.3(R Core Team 2020). I had to download my data through excel, so to get the data into my own data table in r studio I used the packages dplyr (Wickham et. al 2020) and readxl (Wickham and Bryan 2019). I then assigned the function read_excel to the term "download data". This function allows me to read in data coming from an excel file. Within that function I then used the sheet and skip functions to select which sheets of data I was using for my analysis. The data contained many variables I was not interested in, which I removed. The ones I kept were: % of severe housing problems, average number of mentally unhealthy days, % of children in poverty, % fair or poor health, life expectancy, and % smokers. 

## Results

```{r}
library(tidyverse)
library(readxl)
library(corrplot)

# Read Data ---------------------------------------------------------------
ranked_meas_data <- read_excel("Data/2020 County Health Rankings Minnesota Data - v1_0 (3).xlsx", 
                            sheet= "Ranked Measure Data", skip = 1)

ranked_meas_data <- 
  transmute(
    ranked_meas_data,
    FIPS,
    State,
    County,
    Deaths,
    Housing_Probs = `% Severe Housing Problems`,
    FP_Health = `% Fair or Poor Health`,
    Smokers = `% Smokers`,
    Mental_Unhealthy_Days = `Average Number of Mentally Unhealthy Days`,
    Children_Poverty = `% Children in Poverty`
  )

add_meas_data <- read_excel("Data/2020 County Health Rankings Minnesota Data - v1_0 (3).xlsx",
                               sheet = "Additional Measure Data", skip = 1)

add_meas_data <-
  transmute(
    add_meas_data,
    FIPS,
    Life_Expec = `Life Expectancy`
  )

joined_data <- inner_join(ranked_meas_data, add_meas_data, by = "FIPS")


# Plots --------------------------------------------------------------

joined_data %>%
  filter(!is.na(County)) %>% 
  select(-FIPS, -State, -County, -Deaths) %>% 
  cor(use="na.or.complete") %>%
  corrplot(method="circle")

ggplot(joined_data, aes(x=Housing_Probs, y=FP_Health))+ geom_point()+
  geom_smooth(method=lm)

ggplot(joined_data, aes(x=Housing_Probs, y=Life_Expec))+ geom_point()+
  geom_smooth(method=lm)

ggplot(joined_data, aes(x=Mental_Unhealthy_Days, y=FP_Health))+ geom_point()+
  geom_smooth(method=lm)

ggplot(joined_data, aes(x=Mental_Unhealthy_Days, y=Life_Expec))+ geom_point()+
  geom_smooth(method=lm)

ggplot(joined_data, aes(x=Housing_Probs, y=Mental_Unhealthy_Days))+ geom_point()+
  geom_smooth(method=lm)

ggplot(joined_data, aes(x=Children_Poverty, y=FP_Health))+ geom_point()+
  geom_smooth(method=lm)

ggplot(joined_data, aes(x=Children_Poverty, y=Life_Expec))+ geom_point()+
  geom_smooth(method=lm)

ggplot(joined_data, aes(x=Children_Poverty, y=Mental_Unhealthy_Days))+ geom_point()+
  geom_smooth(method=lm)

```

## Discussion



## Literature Cited

Hadley Wickham and Jennifer Bryan (2019). readxl: Read
  Excel Files. R package version 1.3.1.
  https://CRAN.R-project.org/package=readxl

Hadley Wickham, Romain François, Lionel Henry and Kirill
  Müller (2020). dplyr: A Grammar of Data Manipulation. R
  package version 1.0.2.
  https://CRAN.R-project.org/package=dplyr
  
RStudio Team (2020). RStudio: Integrated Development
  Environment for R. RStudio, PBC, Boston, MA URL
  http://www.rstudio.com/.
  
R Core Team (2020). R: A language and environment for
  statistical computing. R Foundation for Statistical
  Computing, Vienna, Austria. URL
  https://www.R-project.org/.

University of Wisconsin Population Health Institute. County Health Rankings 2020.     Available at www.countyhealthrankings.org. Accessed 3/4/2020.
